---
title: "JedR training"
---

Here we introduce and review functions using the [starwars characters](https://dplyr.tidyverse.org/reference/starwars.html) dataset.

The idea is once you are used to the data, you can concentrate on how the functions work.

## Libraries

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
```

## The data

We'll mostly use the `starwars` data already in tidyverse.

Be sure to note films, vehicles & starships. We'll ignore these for the most part.

## JedR: Training Module 1

We'll review some functions and introduce others during this first day.

### Print dataframe

```{r}
starwars
```

### glimpse

Lets you peek at all the data variables at once.

```{r}
starwars |> glimpse()
```

Can also be ...

```{r}
glimpse(starwars)
```


### select

This function selects specific columns or variables. There are many different ways you can specify which columns to keep or drop.

#### select specific variables

We can name which columns to keep:

```{r}
starwars |> select(name, mass)
```

#### select with range

Or a range of columns in order.

```{r}
starwars |> select(name:mass)
```

#### select by order

Or by order of the columns ...

```{r}
starwars |> select(1, 3)
```

#### select with function

There are many special helper functions to grab variables based on their name. Paying attention to how you name your variables can help you take advantage of this:

```{r}
starwars |> select(name, ends_with("color"))
```

#### select to drop

We can drop one variable with ! or -

```{r}
starwars |> select(!name)
```

And even a range of columns

```{r}
starwars |> select(!name:mass)
```

But if the variables are not contiguous (in order together) then we have to use `c()`:

```{r}
starwars |> select(!c(name, mass))
```

### c to concatenate

Concatenate means to join together. `c()` specifies a collection of things.

```{r}
species_short <- 
  c(
    "Human",
    "Droid"
  )

starwars |> 
  filter(species %in% species_short) |> 
  select(name, species)
```

### summary

We use the select here to avoid having the lists because they make the summary display wonky. Usually we would pipe directly into summary.

```{r}
starwars |>
  select(name:species) |> 
  summary()
```

### filter

filter() returns only rows that meet logical criteria you specify.

```{r}
starwars |> 
  filter(species == "Human")
```

```{r}
starwars |> 
  filter(height > 200)
```

```{r}
starwars |> 
  filter(str_detect(hair_color, "auburn"))
```


### mutate

#### Centemeters to inches

multiply by 2.54

```{r}
starwars |> 
  mutate(
    height_in = height * 2.54,
    .after = height
  )
```


## JedR: Training Module 2

### Rounding

The `round()` function in R does not work the way we would expect it to. Instead of rouding up to the next number, it rounds to the nearest even number.

So we use `round_half_up()` that comes with the janitor library.

```{r}
round(11.5)
round(12.5)

round_half_up(11.5)
round_half_up(12.5)
```

### summarize

summarize() builds a summary table about your data. You can count rows n() or do math on numerical values, like mean(). In the next chapter we will summarize with math functions.
 

```{r}
starwars |> 
  summarise(
    avg_height = mean(height, na.rm = TRUE)
  )
```

#### GSA with n

group_by() is often used with summarize() to put data into groups before building a summary table based on the groups.

```{r}
starwars |> 
  group_by(species) |> 
  summarise(numb_chars = n()) |> 
  arrange(desc(numb_chars))
```

#### GSA with sum

```{r}
starwars |> 
  group_by(species) |> 
  summarise(
    numb_chars = n(),
    avg_height = mean(height, na.rm = TRUE) |> round_half_up()
  )
```

### distinct

distinct() returns rows based on unique values in columns you specify. i.e., it deduplicates data.

There are 87 characters in the data, but when we run distinct on eye and hair color, we find there are only 34 combinations of those two variables.

This is not a replacement for group_by or filter or select.

```{r}
starwars |> 
  distinct(hair_color, eye_color)
```

### slice

Slice is sort of like filter, but ... not. It lets you index rows by location.

```{r}
starwars |> 
  slice(1:2)
```

#### slice_sample

But there are many useful tidyverse variations

```{r}
starwars |> 
  slice_sample(n = 4)
```

#### slice_max, slice_min

This allows you to select a row based on value in the data.

```{r}
starwars |> 
  slice_max(mass, n = 2)
```


#### slice with group_by

If you group first, slice will be applied within those groups.

This finds the heaviest person in each species.

```{r}
starwars |> 
  select(name, species, mass) |> 
  group_by(species) |> 
  slice_max(mass) 
```

## JedR: Training Module 3

### nrow

This sometimes comes in handy to save variables for printing in Markdown.

```{r}
starwars |> nrow()

char_rows <- starwars |> nrow()
```

The number of characters in the data: `r char_rows`.


### case tests

Creating simplified variables using case_ functions as tests.

#### case_match

When you look at a specific column/variable and make changes based on those values.

```{r}
starwars |> 
  mutate(
    species_simple = case_match(
      species,
      "Human" ~ "Human",
      .default = "Other"
    )
  ) |> 
  select(name, starts_with("species"))
```

#### case_when

When you need more complicated logic that might look at more than one column (though this doesn't.)

```{r}
starwars |> 
  mutate(
   species_simple = case_when(
     species == "Human" ~ "Human",
     species == "Droid" ~ "Droid",
     .default = "Other"
   )
  ) |> 
  select(name, starts_with("species"))
```

### count

See the special extra chapter in rwdir, but here is an example:

```{r}
starwars |> 
  count(species, sort = T, name = "number_characters")
```


### Pivot

#### pivot wider

These examples are a little harder to contrive with the starwars data, but we'll try.

First, we'll count our characters by species and gender.

```{r}
sw_gen <- starwars |> 
  count(species, gender) |> 
  drop_na(species) # removes rows with no species

sw_gen
```

Now, let's say we want to rearrange this so there is a column for each gender value, one for "masculine", one for "feminine", etc.

We have to specify from which variable to make the new columns, and then from where to grab values.

```{r}
sw_gen_wide <- sw_gen |> 
  pivot_wider(names_from = gender, values_from = n)

sw_gen_wide |> remove_empty()
```

#### pivot longer

Now we flip this longer so there is a row for each species and gender.

```{r}
sw_gen_wide |> 
  pivot_longer(cols = !species, names_to = "gender", values_to = "cnt")
```

## JedR: Training unknown


### separate by delimiter

### separate_longer

[separate longer](https://tidyr.tidyverse.org/reference/separate_longer_delim)

Split the multiple values in skin_color and then count them. The `skin_color` variable starts with values like "white, blue".

```{r}
starwars |> 
  select(species, skin_color) |> 
  separate_longer_delim(skin_color, delim = ",") |> 
  count(species, skin_color)

```

### separate_wider

With [separate wider](https://tidyr.tidyverse.org/reference/separate_wider_delim.html), we are splitting values into new variables.

Again, it is hard to contrive an example with starwars. We have to find a specific set of data to show it without errors.

First we find Droids with two-part names so we can separate by the dash.

```{r}
sw_dash <- starwars |>
  filter(str_detect(name, "-"), species == "Droid") |>
  select(name, species)

sw_dash
```

And now we separate those names into two columns. We use an argument to keep the original column and then move it back to the front.

```{r}
sw_dash |> 
  separate_wider_delim(name, delim = "-", names = c("first", "second"), cols_remove = F) |> 
  relocate(name)
```

## JedR: Training with Rebels

### read

Sometimes `starwars` does not have a data type we need, so we bring in another data set we call `rebels` which is a list of episodes from "Star Wars: Rebels (TV Series 2014-2018)".

We keep a copy of the data at https://bit.ly/jedr-rebels in the [JedR Academy](https://github.com/utdata/jedr-academy) project.

To use it we need to:

- import it
- clean names
- assign to an object

[read_csv](https://readr.tidyverse.org/reference/read_delim.html) and the other readr functions work similarly.

```{r}
rebels <-
  read_csv("https://bit.ly/jedr-rebels") |> 
  clean_names()
```

### Convert a date

Uses dmy (day-month-year or whatever the order) from [lubridate](https://lubridate.tidyverse.org/) to make a real data.

```{r}
rebels |> 
  mutate(
    air_date = dmy(original_air_date),
    .after = original_air_date
  )
```

## JedR: ggplot

Nope ... don't have one.




